Informative Oceanographic Sampling using Excursion Probabilities for Multivariate Random Fields

### 2019-09-30 main_new.tex
Updates Jo

\begin{abstract}
  Improving and optimizing oceanographic sampling is a crucial task
  for modern marine management and science. Faced with limited
  resources to understand highly dynamic processes in the
  water-column, the combination of statistics and autonomous robotics
  provides opportunities for studying aspects related to experimental
  design for effective characterization of oceanographic phenomena. In
  this work we develop methods for efficient spatial sampling applied
  to the mapping of coastal processes by providing informative
  descriptions of spatial characteristics of ocean
  phenomena. Specifically, we define a design criterion based on
  reducing uncertainty in the joint excursions of random fields, and
  derive analytical expressions for the expected Bernoulli variance
  reduction in the excursion set of spatial temperature and salinity
  variables. We show how these can be applicable in static designs and
  sequential designs, building on myopic and look-ahead sampling
  strategies. We use simulations to study the properties and
  differences between state-of-the-art approaches followed by results
  from field deployments with an autonomous underwater vehicle as part
  of a case study mapping the boundary of a river plume. The results
  demonstrate the ability of the method to effectively inform and
  execute data-driven environmental sampling.

@@

\begin{keyword}
\kwd{Ocean Sampling}
\kwd{Excursion Sets}
\kwd{Gaussian Processes}
\kwd{Experimental design}
\kwd{Adaptive information gathering}
\end{keyword}

@@

Traditional data collection at sea has typically been based on static
buoys, lagrangian floats, or ship-based methods, with significant
logistical limitations that directly impact coverage and sampling
resolution. Modern methods using satellite remote-sensing provide
large-scale coverage but have limited resolution, are limited to
sensing the surface of the ocean, and are impacted by cloud cover. The
advent of robust mobile robotic platforms \cite{Bellingham07} has
resulted in significant contributions to environmental monitoring and
sampling. In particular, autonomous underwater vehicles (AUVs), have made
strides in advancing the state of sampling and consequently have made
robotics an integral part of ocean observation; our previous work
has contributed to this effort
% using statistical methods in design and data assimilation
\citep{das11b,Graham2013,das15,Das2015,fossum18b,fossuminformation}.
Others working in this space include \cite{wikle2013modern}, focusing
on hierarchical statistical models; \cite{sahu2008space}, studying
spatio-temporal models for sea surface temperature and salinity data;
and \cite{mellucci2018oceanic} looking at the statistical prediction
of features using an underwater glider. Common to these approaches is
the use of information-based methods to infer environmental sampling
and overcome common challenges, namely:

\begin{itemize}
\item Full numerical ocean models cannot provide accurate results online if run on robotic sensing
  platforms, as on board computers cannot deliver the computational power required,
  hence statistical proxy models of the environment must be used instead.
\item Decisions on where to sample happen sequentially, i.e.  recently
  acquired information can be assimilated into and used to improve
  statistical models and reduce uncertainty associated with where to
  make measurements.
\item As a-priori information about the state of the ocean is very
  uncertain, there is a substantial efficiency gain in reacting to
  information obtained from measurements taken \emph{in-situ}; this is
  usually referred to as the \emph{adaptivity gap}
  \citep{Krause2008phd}.
\item For such types of sampling problems, the number of choices (i.e.
  locations, trajectories, and candidate designs) makes the problem
  combinatorially large, creating a trade-off between optimization
  (finding the most resource-efficient design to collect necessary
  data), and computability (arriving at a solution in reasonable
  time). To successfully resolve features, this trade-off has to be
  considered in sample design and practice.
\end{itemize}

The combination of statistical tools and robotic platforms is a
natural symbiosis which enables information-based sensing. Central to
this is the ability to model spatially-correlated variables and
provide formal measures of uncertainty. In our formulation we used
Gaussian Processes (GPs) in large part because they allow for
efficient implementation and evaluation, in turn enabling
prioritization of sampling efforts which can be be formalized into
decision-making systems on board a robotic platform.

@@

In this work, we build on joint excursion sets (ES) and excursion
probabilities (EP) for the spatial design of experiments. The goal is
to design sampling strategies for reducing the spatially-integrated
Bernoulli variance (IBV) in the ES indicators. Our main contributions
include: i) extending ES-based uncertainty criteria to multivariate
GPs, thereby allowing the evaluation to include multiple environmental
parameters such as temperature and salinity; ii) developing methods
for information-based environmental sensing and autonomous data
collection; and iii) demonstrating the approach in the real-world
scenario of sampling a river plume using an AUV.

The remainder of this paper is organized as follows: Section
\ref{sec:bg} provides select background on ocean sampling; Section
\ref{sec:ESEP} defines ESs, EPs, and the design criteria of IBV;
Section \ref{sec:GP_EP} sets up the required statistical modeling
assumptions of the bi-variate GPs for temperature and salinity;
Section \ref{sec:sur} builds on these assumptions when deriving the
analytical derivation of the design criteria; Section
\ref{sec:heuristics} extends these results and present heuristics for
adaptive sampling; Section \ref{sec:simulations} shows our results and
discusses properties of the methods in simulation studies. Section
\ref{sec:case_study} demonstrates the methodology used in field work
characterizing a river plume; and Section \ref{sec:concl_disc}
contains a summary and a discussion of future work.

@@

Central to understanding the changes taking place in the upper
water-column is an understanding of biophysical interaction driven by
an agglomeration of physical forcing (wind, topography, bathymetry,
tidal influences and local geography, such as fresh water inflows)
and incipient micro-biology driven by planktonic and coastal
anthropogenic input, such as pollution and agricultural runoff. These
often result in a range of ecosystem-related phenomena such as blooms,
anoxic zones and plumes, all of which have significant direct and indirect effects  on society. Walter Munk referred to the
20\textsuperscript{th} century as the \emph{century of undersampling}
\citep{munk2002}, pointing out a lack of sufficient sampling
resolution in time and space as a critical issue for characterizing
ocean phenomena. Even today, policymakers struggle with limited
observability of severe environmental problems such as the recent outbreak of harmful algal blooms in Northern
Norway\footnote{Report from the Institute of Marine
  Research: https://bit.ly/2KZGJfI} (May 2018), with losses exceeding 12 million
tons of salmon valued at \$80 million (Norwegian Directorate of
Fisheries). This is a case in point of how adequate sampling efforts
could have provided a more accurate assessment and prior
warning of changing environmental conditions. These challenges related to undersampling have been driving
ocean sensing practices towards more autonomous, networked, and
distributed use of robotic assets in order to fill the gaps in our
understanding (See Fig. \ref{fig:envir}), where observation is
guided by assimilation of measurements and statistical models.

@@

River plumes belong to a class of ocean processes that are local and
at smaller \emph{sub-mesoscale} (from $5 m^2$ -- $10 km^2$) where
spatial lateral variability tends to dominate. Remote sensing and
synthetic ocean models find it challenging to provide detail at this
scale \citep{Lermusiaux:2006} and the principal way to resolve for
fine resolution is by direct observation. At larger \emph{mesoscale}
($>50 km^2$), both three-dimensional space and time dynamics are
important and can shift substantially, while in the case of river
plumes one can often limit the scope to cover only lateral spatial
elements. Moreover, in this case, time-effects can be regarded as
static when acquiring AUV data over a short time window and limited
region. However, the methodological framework can be extended to
higher dimensional situations in a similar manner.

@@

AUVs are used to gather in-situ spatio-temporal data while carrying a
range of scientific payloads to survey the water column. Typically,
they operate at $\sim 1-3$ m/s, can reach depths from 100--6000 m with
an in-water operation time capacity dependent on survey speed, payload
sensors and mission design. By deploying AUVs together with other
resources, one can augment information obtained from ocean models,
remote-sensing data, or fixed-location sensors on buoys. Traditional
operations and surveys with AUVs are usually limited to observations
along fixed transects on a spatial grid, pre-programmed by the human
operator. By instead using on-board algorithms to continuously
evaluate, update, and refine future sampling locations, one can make
information gathering adaptive and contextual to the survey
requirements \citep{das11b,fossum18b,fossuminformation}. The space of
sampling opportunities is still limited to the spatial grid, but now
there is much more flexibility because of adaptation. For instance, an
AUV could reconstruct or modify a survey line based on what
temperatures it is measuring by using embedded methods in
decision-theoretic planning \citep{py10,Rajan12,Rajan12b}.

@@

Our objective is to leverage statistical tools to characterize a river plume, focusing on spatial separation of cold freshwater from the river and warmer saline waters of the fjord. ES and EPs are useful starting points to measure the ability to classify water masses. The goal of experimental sampling is to improve this characterization, as quantified via changes in the EPs.

We consider a lateral domain near the sea surface, with locations
$\bx \in \mathcal{M} \subset \mathbf{R^2}$. The two variables of
interest, temperature and salinity, are defined as random processes:
$\xi_T(\bx)$ denotes the temperature (in $^o C$) and $\xi_S(\bx)$
denotes the salinity (in $mg/l$) at location $\bx$. The generalization to more than two variables is given in Section \ref{gen_k}.

@@

The ES holds random indicator variables while the associated EPs are: 
\begin{equation}\label{eq:prob}
  p(\bx) = P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S), \hspace{3mm} \bx \in \mathcal{M}.
\end{equation}
If the EPs in Eq. (\ref{eq:prob}) are
close to $1$ or $0$ at a given location, it is easy to classify water mass at this point as either river or ocean; it is, however,
challenging if the probability is close to $0.5$. 
The IBV \cite{bect2019}, is defined as: 
\begin{equation}\label{measVAR}
    V = \int_{\bx \in \mathcal{M}} p(\bx) \left(1-p(\bx)\right) d\bx.
\end{equation}

If prior knowledge provides a strong indication of cold freshwater
 in a part of the domain, it is easy to classify and
there is likely little value in exploring those locations. Data can be
gathered at selected locations of the domain $\mathcal{M}$, to improve
classification performance. We denote temperature measurements by
$y_{T}(\bx_i)$ and salinity measurements by $y_{S}(\bx_i)$ at a
sampling location $\bx_i \in \mathcal{M}$. Given a set of
measurements denoted $\by$, the conditional EPs are:
\begin{equation}\label{eq:post_ep}
 p(\bx;\by) = P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S |\by). 
\end{equation}

Our goal is to construct adaptive sampling strategies based on continuous evaluation of the EPs. As the EPs provide a description of the boundary uncertainty, the strategies can effectively use this description to prioritize sampling efforts at locations that are ambiguous, making exploration more effective. As the AUV gathers data throughout its mission, it is natural to evaluate the expected reduction in the IBV, conditioned on the survey data. In this way, we can
take the expectation with respect to the planned measurements $\by$
\begin{equation}\label{sur}
    V_{\mbox{upd}} = \int_{\bx \in \mathcal{M}} E_{\by} \left\{ p(\bx;\by)\left( 1-p(\bx;\by)\right) \right\} d\bx, 
\end{equation}
where the expectation is over the planned data, and the probability is defined in Eq. (\ref{eq:post_ep}).
Considering a set $\mathcal{J}$ of possible designs for an AUV survey, a criterion for the selection is
\begin{equation}\label{crit}
    j^* = \mbox{argmin}_{j \in \mathcal{J}} V_{\mbox{upd}}(j),
\end{equation}
where the criterion $V_{\mbox{upd}}(j)$ in Eq. (\ref{sur}) is computed
for each of the possible designs.

Recent work on ESs and EPs connected to spatial statistics include
\cite{picheny2010,french2013spatio,bolin2015excursion,french2016credible}.
Our focus is on ES as defined in Eq. (\ref{ES}) and the uncertainty
reduction achieved by sampling as in Eq. (\ref{sur}). In this sense,
our work is similar to \cite{bect2012}, \cite{chevalier2014fast} and
\cite{azzimonti2016quantifying} who describe analytical results for
uncertainty reduction in ESs for univariate processes. A major
contribution of our work is to derive closed-form results for this
design criteria for situations where the underlying model is based on multivariate GPs.

@@

Calculations of the EPs and the IBV require a model specification. We use Gaussian modeling, which provides a practical and popular probabilistic approach to modeling environmental data, and allows a closed form
evaluation of the IBVs as derived in Section \ref{sec:sur}.

\subsection{Bivariate Gaussian Processes}

The Gaussian model is used to represent the bivariate random fields of temperature and salinity. At location $\bx \in \mathcal{M}$, the prior bivariate distribution is
\begin{equation}\label{gp_bar}
  \begin{bmatrix}\xi_T(\bx) \\
    \xi_S(\bx) \end{bmatrix}
 \sim N \left( 
\begin{bmatrix} \mu_{T}(\bx)\\
\mu_{S}(\bx)
\end{bmatrix},\begin{bmatrix}
\sigma_{T}^2(\bx) & \sigma_{T}(\bx) \sigma_S(\bx) \gamma(\bx)  \\
\sigma_{T}(\bx) \sigma_S(\bx) \gamma(\bx)  & \sigma_{S}^2(\bx) 
\end{bmatrix}
\right),
\end{equation}
where the notation refers to Gaussian (i.e. normal) distributed variables
specified by the mean vector and covariance matrix. For the application
of tracking river plumes, the model parameters are specified from a short preliminary survey. The mean varies with location, capturing the variability near the river mouth, while the variance parameters and correlation between salinity and temperature are assumed to be constant for all locations and within each water mass.

For the spatial dependence among variables, we assume a separable correlation function with the same decay for salinity and temperature,
which is consistent with expectations for  a river plume where the decay is related to the common water masses (Sect \ref{sec:case_study}). In the resulting function, we have
\begin{equation}\label{gp_corr}
\mbox{Cov}(\xi_i(\bx),\xi_j(\bx')) = \rho(\bx,\bx') \sigma_i \sigma_j (\gamma +(1-\gamma )\delta_{ij}), \\
\end{equation}
where for $i,j \in {T,S}$ and $\delta_{ij}=1$ if $i=j$, and $\delta_{ij}=0$
otherwise. Moreover, we assume stationary isotropic random fields a priori where the
correlation function $\rho(\bx,\bx')$ solely depends on $\bx$ and
$\bx'$ via a Euclidean distance measure $h=\sqrt{||\bx-\bx'||^2}$. With
data and prior knowledge, one could possibly fit and estimate
parameters of non-stationary or non-isotropic covariance functions and
more complex multivariate spatial covariance functions
\citep{gneiting2010matern,genton2015cross}; however, that is left for
future work.

We discretize the spatial domain
$\mathcal{M}$ to a set of $n$ grid locations
$\mathcal{M}_g = \{\bx_i, i=1,\ldots,n \}$, where each cell has area $\Delta$. The
grid is used for the waypoint graph setting the possible design locations. It is also used when solving integrals expressions in Eq. (\ref{measVAR}) and
(\ref{sur}), which are then approximated by sums over all grid cells.  Using vector
notation, we denote the Gaussian temperature and salinity variables on
the grid and its density function by the formula: 
\begin{equation}\label{prior}
    \bxi = (\xi_T(\bx_1),\xi_S(\bx_1),\ldots,\xi_T(\bx_n),\xi_S(\bx_n))^t, \hspace{3mm}
    \bxi  \sim  N(\bmu, \bSigma) %\pi (\bxi) & =& \frac{1}{(2\pi)^{n}|\bSigma|^{\frac{1}{2}}}e^{-\frac{1}{2}(\bxi -\bmu)^T\bSigma^{-1}(\bxi-\bmu)}.
\end{equation}
%Here, $\pi(\bxi)$ is the joint density function of length $2 n$ vector $\bxi$, and
The length $2 n$ vector $\bmu$ contains the expected values of
temperature and salinity at the grid locations, while the
$2n \times 2n$ matrix $\bSigma$ contains the
variance-covariances of temperature and salinity variables at all grid
locations.

\subsection{Data Model and Updating of Gaussian Processes}

Data is gathered by an AUV that can measure temperature and salinity
at a subset of locations on the defined grid. This can be done in a
static manner (where the survey design is pre-planned), or in a
sequential way with several rounds of data gathering. For the latter,
a round of data is gathered first, and then assimilated to get an
updated model. Then the updated model is used to design and select the
next round of measurements, and so on. Below, we describe the model for one round of data and the Gaussian formula for a single update; in Section \ref{sec:heuristics}, we discuss sequential data gathering in the context of adaptive design of AUV measurements.

@@

The updated bivariate distribution at a grid location $\bx \in
\mathcal{M}_g$ is then: 
\begin{equation}\label{gp_hat}
\begin{bmatrix}
\xi_T(\bx) \\
\xi_S(\bx)
\end{bmatrix}
 |\by
 \sim N \left( 
\begin{bmatrix} m_{T}(\bx)\\
m_{S}(\bx)
\end{bmatrix},\begin{bmatrix}
a_{T}^2(\bx) & a_{T,S}(\bx)  \\
a_{T,S}(\bx)  & a_{S}^2(\bx)  
\end{bmatrix}
\right),
\end{equation}
with entries $m_{\cdot}$ and $a_{\cdot}$ extracted from the conditional mean and covariance
expressions in Eq. (\ref{gp_upd}).

% The conditional mean $\bm$ will be important in the following
% derivation in Section \ref{sec:sur}.
The calculation of the design criterion in Eq. (\ref{sur}) requires
the marginal distribution of the data which is defined over $\bxi$ to
be $\by \sim N( \bG \bmu,\bG \bSigma \bG^t+\bR)$.  However, it turns
out that a simplified form of the calculation is possible because the
conditional mean in Eq (\ref{gp_hat}) is a linear (affine) function of
the data $\by$ and the covariance is not a function of the data.
Before inputting the data, the distribution of the conditional mean is
\begin{equation}\label{distxi} \bm \sim N(\bmu , \bPsi), \hspace{3mm}
  \bPsi=\bSigma \bG^t (\bG \bSigma \bG^t+\bR)^{-1} \bG \bSigma.
\end{equation} 
When considering only grid location $\bx$, we can
extract elements from its mean vector and covariance matrix to get the
bivariate distribution \begin{equation}\label{dist_mxi} 
  \bm(\bx) \sim N \left( \begin{bmatrix}
      \mu_{T}(\bx) \\
      \mu_{S}(\bx) \end{bmatrix}, \begin{bmatrix}
      \psi^2_{T}(\bx) & \psi_{T,S}(\bx)\\
      \psi_{T,S}(\bx) & \psi^2_{S}(\bx) \end{bmatrix} \right),
\end{equation}
which is important for the derivation of the closed-form solution of the design criteria.

@@

\subsection{Excursion Probabilities for Gaussian Processes}

>>>>

\begin{abstract}
  Improving and optimizing oceanographic sampling is a crucial task
  for modern marine management and science. Faced with limited
  resources to understand highly processes in the
  water-column, the combination of statistics and autonomous robotics
  provides opportunities for studying aspects related to experimental
  design for effective characterization of oceanographic phenomena. In
  this work we develop methods for efficient spatial sampling applied
  to the mapping of coastal processes by providing informative
  descriptions of spatial characteristics of ocean
  phenomena. Specifically, we define a design criterion based on
  reducing uncertainty in the joint excursions of random fields, and
  derive analytical expressions for the expected Bernoulli variance
  reduction in the excursion set of spatial temperature and salinity
  variables. We show how these can be applicable in static designs and
  sequential designs, building on myopic and look-ahead sampling
  strategies. We use simulations to study the properties and
  differences between state-of-the-art approaches followed by results
  from field deployments with an autonomous underwater vehicle as part
  of a case study mapping the boundary of a river plume. The results
  demonstrate the ability of the method to effectively inform and
  execute data-driven environmental sampling.

@@

\begin{keyword}
\kwd{Ocean Sampling}
\kwd{Excursion Sets}
\kwd{Gaussian Processes}
\kwd{Experimental Design}
\kwd{Adaptive Information Gathering}
\end{keyword}

@@

Traditional data collection at sea has typically been based on static
buoys, Lagrangian floats, or ship-based methods, with significant
logistical limitations that directly impact coverage and sampling
resolution. Modern methods using satellite remote-sensing provide
large-scale coverage but have limited resolution, are limited to
sensing the surface of the ocean, and are impacted by cloud cover. The
advent of robust mobile robotic platforms \citep{Bellingham07} has
resulted in significant contributions to environmental monitoring and
sampling. In particular, autonomous underwater vehicles (AUVs), have advanced the state of sampling and consequently have made
robotics an integral part of ocean observation; our previous work
has contributed to this effort
% using statistical methods in design and data assimilation
\citep{das11b,Das2015,fossum18b,fossuminformation}.
Other statistical work in the oceanographic domain include \cite{wikle2013modern} focusing
on hierarchical statistical models; \cite{sahu2008space}, studying
spatio-temporal models for sea surface temperature and salinity data;
and \cite{mellucci2018oceanic} looking at the statistical prediction
of features using an underwater glider.

The goal of the information-based methods that we highlight in the current paper, is to do efficient environmental sampling and overcome common challenges, namely:

\begin{itemize}
\item Full numerical ocean models cannot provide accurate results online if run on robotic sensing
  platforms, as onboard computers cannot deliver the computational power required,
  hence statistical proxy models of the environment must be used instead.
\item Decisions on where to sample happen sequentially, i.e. recently
  acquired information can be assimilated into and used to improve
  statistical models and reduce uncertainty associated with where to
  make measurements.
\item As a priori information about the state of the ocean is rather
  uncertain, there is a substantial efficiency gain in reacting to
  information obtained from measurements taken in-situ. %\emph{in-situ}; this is usually referred to as the \emph{adaptivity gap} \citep{Krause2008phd}.
\item For such types of sampling problems, the number of choices (i.e.
  locations, trajectories, and candidate designs) makes the problem
  combinatorially large, creating a trade-off between optimization
  (finding the most resource-efficient design to collect necessary
  data), and computability (arriving at a solution in reasonable
  time). To successfully resolve features, this trade-off has to be
  considered in sample design and practice.
\end{itemize}

The combination of statistical tools and robotic platforms is a
natural symbiosis which enables information-based sensing. Central to
this is the ability to model spatially-correlated variables and
provide formal measures of uncertainty. In our formulation we use
Gaussian Processes (GPs) in large part because they allow for
efficient implementation and evaluation, in turn enabling
prioritization of sampling efforts which can be formalized into decision-making systems onboard a robotic platform.

@@

In this work, we build on joint excursion sets (ES) and excursion
probabilities (EP) for the spatial design of experiments. The goal is
to design sampling strategies for reducing the spatially-integrated
Bernoulli variance (IBV) in the ES indicators. Our main contributions
include: i) extending the IBV criteria to multivariate
GPs, thereby allowing the evaluation to include multiple environmental
parameters such as temperature and salinity; ii) developing methods
for information-based environmental sensing and autonomous data
collection; and iii) demonstrating the approach in the real-world
scenario of sampling a river plume using an AUV.

The remainder of this paper is organized as follows: Section
\ref{sec:bg} provides select background on ocean sampling; Section
\ref{sec:ESEP} defines ESs, EPs, and the design criteria of IBV;
Section \ref{sec:GP_EP} sets up the required statistical modeling
assumptions of the bi-variate GPs for temperature and salinity;
Section \ref{sec:sur} builds on these assumptions when deriving the
analytical derivation of the design criteria; Section
\ref{sec:heuristics} extends these results and present heuristics for
adaptive sampling; Section \ref{sec:simulations}
discusses properties of the methods in simulation studies. Section
\ref{sec:case_study} demonstrates the methodology used in field work
characterizing a river plume; and Section \ref{sec:concl_disc}
contains a summary and a discussion of future work.

@@

Central to understanding the changes taking place in the upper
water-column is an understanding of biophysical interaction driven by
an agglomeration of physical forcing (wind, topography, bathymetry,
tidal influences and local geography, such as fresh water inflows)
and incipient micro-biology driven by planktonic and coastal
anthropogenic input, such as pollution and agricultural runoff. These
often result in a range of ecosystem-related phenomena such as blooms,
anoxic zones and plumes, all of which have significant direct and indirect effects  on society. Walter Munk referred to the
20\textsuperscript{th} century as the \emph{century of undersampling}
\citep{munk2002}, pointing out a lack of sufficient sampling
resolution in time and space as a critical issue for characterizing
ocean phenomena. Even today, policymakers struggle with limited
observability of severe environmental problems such as the recent outbreak of harmful algal blooms in Northern
Norway\footnote{Report from the Institute of Marine
  Research: https://bit.ly/2KZGJfI} (May 2018), with losses exceeding 12 million
tons of salmon valued at \$80 million (Norwegian Directorate of
Fisheries). This is a case in point of how adequate sampling efforts
could have provided a more accurate assessment and early
warning of changing environmental conditions. These challenges related to undersampling have been driving
ocean sensing practices towards more autonomous, networked, and
distributed use of robotic assets in order to fill the gaps in our
understanding (See Fig. \ref{fig:envir}), where observation is
guided by assimilation of measurements and statistical models.

@@

River plumes belong to a class of ocean processes that are local and
at smaller \emph{sub-mesoscale} (from $5 m^2$ -- $10 km^2$) where
spatial lateral variability tends to dominate. Remote sensing and
synthetic ocean models find it challenging to provide detail at this
scale \citep{Lermusiaux:2006} and the principal way to resolve for
fine resolution is by direct observation. At larger \emph{mesoscale}
($>50 km^2$), both three-dimensional space and time dynamics are
important and can shift substantially, while in the case of river
plumes one can often limit the scope to cover only lateral spatial
elements. However, the methodological framework can be extended to
higher dimensional situations in a similar manner.

@@

AUVs are used to gather in-situ spatio-temporal data while carrying a
range of scientific payloads to survey the water column. Typically,
they operate at $\sim 1-3$ m/s, can reach depths from 100--6000 m with
an in-water operation time capacity dependent on survey speed, payload
sensors and mission design. By deploying AUVs together with other
resources, one can augment information obtained from output from synthetic ocean models,
remote-sensing data, or fixed-location sensors on buoys. Traditional
operations and surveys with AUVs are usually limited to observations
along fixed transects on a spatial grid, pre-programmed by the human
operator. By instead using onboard algorithms to continuously
evaluate, update, and refine future sampling locations, one can make
information gathering adaptive and contextual to the survey
requirements \citep{das11b,fossum18b,fossuminformation}. The space of
sampling opportunities is still limited to the spatial grid, but now
there is much more flexibility because of adaptation. For instance, an
AUV could reconstruct or modify a survey line based on what
temperatures it is measuring by using embedded methods in
decision-theoretic planning \citep{py10,Rajan12,Rajan12b}.

@@

Our objective is to leverage statistical tools to characterize a river plume, focusing on spatial separation of cold freshwater from the river and warmer saline waters of the fjord. ES and EPs are useful starting points to measure the ability to classify water masses. The goal of experimental sampling is to improve this characterization. %, as quantified via changes in the EPs.

We consider a lateral domain near the sea surface, with locations
$\bx \in \mathcal{M} \subset \mathbf{R^2}$. The two variables of
interest, temperature and salinity, are defined as random processes:
$\xi_T(\bx)$ denotes the temperature (in $^o C$) and $\xi_S(\bx)$
denotes the salinity (in $mg/l$) at location $\bx$. The generalization to more than two variables is given in Section \ref{gen_k}. The adaptive sampling strategies are described in Section \ref{sec:heuristics}.

@@

The ES holds random indicator variables while the associated EPs are: 
\begin{equation}\label{eq:prob}
  p(\bx) = P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S), \hspace{3mm} \bx \in \mathcal{M}.
\end{equation}
If the EPs in Eq. (\ref{eq:prob}) are
close to $1$ or $0$ at a given location, it is easy to classify water mass at this point as either river or ocean; it is, however,
challenging if the probability is close to $0.5$. 
The IBV, see \cite{bect2019}, is defined as: 
\begin{equation}\label{measVAR}
    V = \int_{\bx \in \mathcal{M}} p(\bx) \left(1-p(\bx)\right) d\bx.
\end{equation}

Data can be
gathered at selected locations of the domain $\mathcal{M}$, to improve
classification performance. We denote temperature measurements by
$y_{T}(\bx_i)$ and salinity measurements by $y_{S}(\bx_i)$ at a
sampling location $\bx_i \in \mathcal{M}$. Given a set of
measurements denoted $\by$, the conditional EPs are:
\begin{equation}\label{eq:post_ep}
 p(\bx;\by) = P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S |\by). 
\end{equation}

Our goal is to construct adaptive sampling strategies based on continuous evaluation of the EPs. As the EPs provide a description of the boundary uncertainty, the strategies can effectively use this description to prioritize sampling efforts at locations that are ambiguous, making exploration more effective. It is natural to evaluate the expected reduction in the IBV, conditioned on the survey data. The expected updated IBV is
\begin{equation}\label{sur}
    V_{\mbox{upd}} = \int_{\bx \in \mathcal{M}} E_{\by} \left( p(\bx;\by)\left( 1-p(\bx;\by)\right) \right) d\bx, 
\end{equation}
where the expectation is over the planned data, and the probability is defined in Eq. (\ref{eq:post_ep}).
Considering a set $\mathcal{J}$ of possible designs for an AUV survey, a criterion for the selection is
\begin{equation}\label{crit}
    j^* = \mbox{argmin}_{j \in \mathcal{J}} V_{\mbox{upd}}(j),
\end{equation}
where the criterion $V_{\mbox{upd}}(j)$ in Eq. (\ref{sur}) is computed
for each of the possible designs.

Recent work on ESs and EPs connected to spatial statistics include
\cite{picheny2010,french2013spatio,bolin2015excursion,french2016credible}.
Our focus is on ES as defined in Eq. (\ref{ES}) and the uncertainty
reduction achieved by sampling as in Eq. (\ref{sur}). In this sense,
our work is similar to \cite{bect2012}, \cite{chevalier2014fast} and
\cite{azzimonti2016quantifying} who describe analytical results for the IBV for univariate processes. A major
contribution of our work is to derive closed-form results for this
design criteria for situations where the underlying model is based on multivariate GPs.

@@

Calculations of the EPs and the IBV require a model specification. We use GP models, which provides a practical and popular probabilistic approach to modeling environmental data, and allows a closed form
evaluation of the IBVs as derived in Section \ref{sec:sur}. GPs can also be fitted to run onboard robotic vehicles with limited computational assets, such as AUVs.

\subsection{Bivariate GPs}

The Gaussian model is used to represent the bivariate random fields of temperature and salinity. At location $\bx \in \mathcal{M}$, the prior bivariate distribution is
\begin{equation}\label{gp_bar}
  \begin{bmatrix}\xi_T(\bx) \\
    \xi_S(\bx) \end{bmatrix}
 \sim N \left( 
\begin{bmatrix} \mu_{T}(\bx)\\
\mu_{S}(\bx)
\end{bmatrix},\begin{bmatrix}
\sigma_{T}^2(\bx) & \sigma_{T}(\bx) \sigma_S(\bx) \gamma(\bx)  \\
\sigma_{T}(\bx) \sigma_S(\bx) \gamma(\bx)  & \sigma_{S}^2(\bx) 
\end{bmatrix}
\right),
\end{equation}
where the notation refers to Gaussian (i.e. normal) distributed variables
specified by the mean vector and covariance matrix. For the application
of a river plume, the model parameters are specified from a short preliminary survey. The means vary with location, capturing the variability near the river mouth, while the variance parameters and correlation between salinity and temperature are assumed to be constant for all locations. %and within each water mass.

For the spatial dependence among variables, we assume a separable correlation function with the same decay for salinity and temperature,
which is consistent with expectations for  a river plume where the decay is related to the common water masses (Section \ref{sec:case_study}). In the resulting function, we have
\begin{equation}\label{gp_corr}
\mbox{Cov}(\xi_i(\bx),\xi_j(\bx')) = \rho(\bx,\bx') \sigma_i \sigma_j (\gamma +(1-\gamma )\delta_{ij}), \\
\end{equation}
where for $i,j \in {T,S}$ and $\delta_{ij}=1$ if $i=j$, and $\delta_{ij}=0$
otherwise. Moreover, we assume stationary isotropic random fields a priori where the
correlation function $\rho(\bx,\bx')$ solely depends on $\bx$ and
$\bx'$ via Euclidean distance $h=\sqrt{||\bx-\bx'||^2}$. With
data and prior knowledge, one could possibly fit and estimate
parameters of non-stationary or non-isotropic covariance functions and
more complex multivariate spatial covariance functions
\citep{gneiting2010matern,genton2015cross}; however, that is left for
future work.

We discretize the spatial domain
$\mathcal{M}$ to a set of $n$ grid locations
$\mathcal{M}_g = \{\bx_i, i=1,\ldots,n \}$, where each cell has area $\Delta$. The
grid is used for the waypoint graph setting the possible design locations. It is also used when solving integral expressions in Eq. (\ref{measVAR}) and
(\ref{sur}), which are then approximated by sums over all grid cells.  Using vector
notation, we denote the Gaussian temperature and salinity variables on
the grid and its density function by the formula: 
\begin{equation}\label{prior}
    \bxi = (\xi_T(\bx_1),\xi_S(\bx_1),\ldots,\xi_T(\bx_n),\xi_S(\bx_n))^t, \hspace{3mm}
    \bxi  \sim  N(\bmu, \bSigma). %\pi (\bxi) & =& \frac{1}{(2\pi)^{n}|\bSigma|^{\frac{1}{2}}}e^{-\frac{1}{2}(\bxi -\bmu)^T\bSigma^{-1}(\bxi-\bmu)}.
\end{equation}
%Here, $\pi(\bxi)$ is the joint density function of length $2 n$ vector $\bxi$, and
The length $2 n$ vector $\bmu$ contains the expected values of
temperature and salinity at the grid locations, while the
$2n \times 2n$ matrix $\bSigma$ contains the
variance-covariances of temperature and salinity variables at all grid
locations.

\subsection{Data Model and Updating of GPs}

Data are gathered by an AUV that can measure temperature and salinity
at a subset of locations on the defined grid. This can be done in a
static manner (where the survey design is pre-planned), or in a
sequential way with several rounds of data gathering. For the latter,
a round of data is gathered first, and then assimilated to get an
updated model. Then the updated model is used to design and select the
next round of measurements, and so on (Section \ref{sec:heuristics}). Below, we describe the model for one round of data and the Gaussian formula for a single update.

@@

The updated bivariate distribution at a grid location $\bx \in
\mathcal{M}_g$ is then: 
\begin{equation}\label{gp_hat}
\begin{bmatrix}
\xi_T(\bx) \\
\xi_S(\bx)
\end{bmatrix}
 |\by
 \sim N \left( 
\begin{bmatrix} m_{T}(\bx)\\
m_{S}(\bx)
\end{bmatrix},\begin{bmatrix}
a_{T}^2(\bx) & a_{T,S}(\bx)  \\
a_{T,S}(\bx)  & a_{S}^2(\bx)  
\end{bmatrix}
\right),
\end{equation}
with entries extracted from the conditional mean and covariance
expressions in Eq. (\ref{gp_upd}).

% The conditional mean $\bm$ will be important in the following
% derivation in Section \ref{sec:sur}.
The calculation of the design criterion in Eq. (\ref{sur}) requires
the marginal distribution of the data which is defined over $\bxi$ to
be $\by \sim N( \bG \bmu,\bG \bSigma \bG^t+\bR)$.  However, it turns
out that a simplified form of the calculation is possible because the
conditional mean in Eq. (\ref{gp_hat}) is a linear (affine) function of
the data $\by$ and the covariance is not a function of the data.
Before inputting the data, the distribution of the conditional mean is
\begin{equation}\label{distxi} \bm \sim N(\bmu , \bPsi), \hspace{3mm}
  \bPsi=\bSigma \bG^t (\bG \bSigma \bG^t+\bR)^{-1} \bG \bSigma.
\end{equation} 
When considering only grid location $\bx$, we can
extract elements from its mean vector and covariance matrix to get the
bivariate distribution \begin{equation}\label{dist_mxi} 
  \bm(\bx) \sim N \left( \begin{bmatrix}
      \mu_{T}(\bx) \\
      \mu_{S}(\bx) \end{bmatrix}, \begin{bmatrix}
      \psi^2_{T}(\bx) & \psi_{T,S}(\bx)\\
      \psi_{T,S}(\bx) & \psi^2_{S}(\bx) \end{bmatrix} \right),
\end{equation}
which is important for the derivation of the closed-form solution of the design criteria.

@@

\subsection{EPs for GPs}
